Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2021-09-29 vertical-thermal-review.Rmd
rename gas exchange section (plus minor edits)

Here, we review how the biophysical environment and plant form and function vary across the vertical canopy gradient in forests. 
We focus on five key themes (Fig. `r fig_schematic`):
(1) the biophysical environment;
(2) leaf temperature ($T_{leaf}$);
(3) the leaf traits that most strongly influence $T_{leaf}$;
(4) leaf metabolism and gas exchange, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
(5) tree and ecosystem ecology.
We then consider the implications for understanding forest responses to global change, including how these responses scale across space and time.

@@

Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor, with the profile shape modified by leaf area density, canopy height, and canopy structure across species and forest types [Fig. `r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001].
This gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`d), than in open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Fig. `r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. 
Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012]. 
The implication of this light gradient is that upper canopy leaves have higher potential photosynthetic rate (see section "Leaf metabolism and thermal responses"), but also greater propensity towards elevated $T_{leaf}$ (see section "Leaf temperature").

@@

$T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other micrometeorological drivers (reviewed above) and by leaf traits and stomatal conductance (see "Leaf traits" and "Leaf metabolism and thermal responses"), with leaves adjusting to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020]. 
Leaves are typically cooler than the air at night, and under some daytime conditions (cloudy skies, high wind speeds, and high $T_{air}$). 
Leaves are typically warmer than air when under full sunlight, especially under slow wind speeds and low $T_{air}$ [@doughtyAreTropicalForests2008].
While $T_{leaf}$ is rarely exactly equal to $T_{air}$, it is most commonly within a few degrees [@cavaleriColdbloodedForestsWarming2020].

@@

The basic biophysical principles outlined above shape the tendency for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in closed-canopy forests.
High radiation at the top of the vertical profile (Fig. `r fig_NEON_vertical`) implies that canopy leaves have much greater tendency for high $T_{leaf}-T_{air}$, such that they would be expected to be warmer than understory leaves under most conditions (Fig. `r fig_leaf_T`).
However, higher wind speeds (Fig. `r fig_NEON_vertical`) reduce $T_{leaf}-T_{air}$ [Fig. `r fig_leaf_T`b, @bonan_ecological_2016; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. 
In addition, adaptive leaf traits that increase $g_b$ and $g_s$ (see "Leaf Traits" and "Leaf metabolism and thermal responses" sections) mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig. `r fig_leaf_T`b-e). 
In contrast, greater RH and lower wind speeds in the understory may limit $\lambda E$ (through reduced $g_s$ and $g_b$), thereby increasing $T_{leaf}$ [Fig. `r fig_leaf_T`, @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979b].

@@

Sun leaves have anatomical, morphological, and physiological traits that reduce heat due to higher conductance ($g_b$ or maximal $g_s$) and/or reflectance, which help to lower $T_{leaf}$-$T_{air}$ (Table `r table_leaf_traits`). 
Thus, sun leaves are generally smaller, thicker, with higher leaf mass per area, and are more deeply lobed [Fig. `r fig_leaf_T`, @leighInfluenceLeafSize2017; @vogelSunLeavesShade1968a; @sackHowStrongIntracanopy2006;@zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a], but with greater leaf packing and clumping (reduces $g_b$).
Steeper leaf angles reduce radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], while higher trichome density increases reflectance, thereby also decreasing radiation load.
Further, sun leaves tend to have higher stomatal and vein densities, which enable higher maximal $g_s$ and thereby facilitate effective cooling [see also "Leaf metabolism and thermal responses" section, @zwienieckiHydraulicLimitationsImposed2004]. 
At the same time, sun leaves also have adaptations to protect against water stress, particularly in drier climates, including greater cuticle thickness (protects against water loss) and higher trichome density (reduces $g_b$; **REF?**).
In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA and higher light absorptance efficiency per unit biomass), but their greater thermal capacitance due to larger leaf size and lower transpiration makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b]. 
In open canopies, where light is comparatively homogeneous, leaf traits may be more shaped by maximum air temperature ($T_{air_{max}}$) and VPD stress than to light [@mediavillaFoliarPlasticityRelated2019a].

Leaf biochemical traits also vary strongly across light and height gradients (Table `r table_leaf_traits`), both enabling and resulting from gradients in metabolism (see "Leaf metabolism and thermal responses" section, Table `r table_leaf_metabolism`).
Sun leaves have higher concentrations (per unit area) of elements such as nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical to leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes [@scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014].
Increases in chlorophyll a/b ratios with height reflect the greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights within the canopy give shade leaves greater PAR absorptance efficiency (Table `r table_leaf_traits`).
Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure higher (see "Leaf metabolism and thermal responses" section, Table `r table_leaf_metabolism`) result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C concentrations in leaf tissues (Table `r table_leaf_traits`).

@@

## Leaf metabolism and thermal responses

@@

\newpage
**Table `r table_leaf_metabolism`.** Summary of observed variation in leaf metabolism and thermal responses across the vertical gradient and/or between sun and shade leaves. Studies listed here were compiled using a systematic review process, as described in Supplementary Information Methods `r appendix_lit_review_methods`.

>>>>

Here, we review how the biophysical environment and plant form and function vary across the vertical canopy gradient in forests. 
We focus on five key themes (Fig. `r fig_schematic`):
(1) the biophysical environment;
(2) leaf temperature ($T_{leaf}$);
(3) the leaf traits that most strongly influence $T_{leaf}$;
(4) leaf gas exchange, including stomatal and boundary layer conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
(5) tree and ecosystem ecology.
We then consider the implications for understanding forest responses to global change, including how these responses scale across space and time.

@@

Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor, with the profile shape modified by leaf area density, canopy height, and canopy structure across species and forest types [Fig. `r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001].
This gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`d), than in open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Fig. `r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. 
Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012]. 
The implication of this light gradient is that upper canopy leaves have higher potential photosynthetic rate (see section "Leaf gas exchange and its thermal sensitivity"), but also greater propensity towards elevated $T_{leaf}$ (see section "Leaf temperature").

@@

$T_{leaf}$ is strongly tied to $T_{air}$, and is further influenced by other micrometeorological drivers (reviewed above) and by leaf traits and stomatal conductance (see "Leaf traits" and "Leaf gas exchange and its thermal sensitivity"), with leaves adjusting to their environment to approach, when possible, an optimal $T_{leaf}$ for carbon assimilation and metabolic processes [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020; @drakeNoEvidenceHomeostatic2020]. 
Leaves are typically cooler than the air at night, and under some daytime conditions (cloudy skies, high wind speeds, and high $T_{air}$). 
Leaves are typically warmer than air when under full sunlight, especially under slow wind speeds and low $T_{air}$ [@doughtyAreTropicalForests2008].
While $T_{leaf}$ is rarely exactly equal to $T_{air}$, it is most commonly within a few degrees [@cavaleriColdbloodedForestsWarming2020].

@@

The basic biophysical principles outlined above shape the tendency for $T_{leaf}$ to be elevated above $T_{air}$ across vertical gradients in closed-canopy forests.
High radiation at the top of the vertical profile (Fig. `r fig_NEON_vertical`) implies that canopy leaves have much greater tendency for high $T_{leaf}-T_{air}$, such that they would be expected to be warmer than understory leaves under most conditions (Fig. `r fig_leaf_T`).
However, higher wind speeds (Fig. `r fig_NEON_vertical`) reduce $T_{leaf}-T_{air}$ [Fig. `r fig_leaf_T`b, @bonan_ecological_2016; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. 
In addition, adaptive leaf traits that increase $g_b$ and $g_s$ (see "Leaf Traits" and "Leaf gas exchange and its thermal sensitivity" sections) mediate the direct effects of solar radiation on upper canopy leaves and result in higher rates of heat loss (Fig. `r fig_leaf_T`b-e). 
In contrast, greater RH and lower wind speeds in the understory may limit $\lambda E$ (through reduced $g_s$ and $g_b$), thereby increasing $T_{leaf}$ [Fig. `r fig_leaf_T`, @perezIncreasingHumidityThreatens2018; @tibbittsHumidityPlants1979b].

@@

Sun leaves have anatomical, morphological, and physiological traits that reduce heat due to higher conductance ($g_b$ or maximal $g_s$) and/or reflectance, which help to lower $T_{leaf}$-$T_{air}$ (Table `r table_leaf_traits`). 
Thus, sun leaves are generally smaller, thicker, with higher leaf mass per area, and are more deeply lobed [Fig. `r fig_leaf_T`, @leighInfluenceLeafSize2017; @vogelSunLeavesShade1968a; @sackHowStrongIntracanopy2006;@zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a], but with greater leaf packing and clumping (reduces $g_b$).
Steeper leaf angles reduce radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], while higher trichome density increases reflectance, thereby also decreasing radiation load.
Further, sun leaves tend to have higher stomatal and vein densities, which enable higher maximal $g_s$ and thereby facilitate effective cooling [see also "Leaf gas exchange and its thermal sensitivity" section, @zwienieckiHydraulicLimitationsImposed2004]. 
At the same time, sun leaves also have adaptations to protect against water stress, particularly in drier climates, including greater cuticle thickness (protects against water loss) and higher trichome density (reduces $g_b$; **REF?**).
In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA and higher light absorptance efficiency per unit biomass), but their greater thermal capacitance due to larger leaf size and lower transpiration makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b]. 
In open canopies, where light is comparatively homogeneous, leaf traits may be more shaped by maximum air temperature ($T_{air_{max}}$) and VPD stress than by light [@mediavillaFoliarPlasticityRelated2019a].

Most leaf biochemical traits also vary across light and height gradients (Table `r table_leaf_traits`), both shaping and resulting from gradients in metabolism (see "Leaf gas exchange and its thermal sensitivity" section, Table `r table_leaf_metabolism`).
Sun leaves have higher concentrations (per unit area) of elements such as nitrogen ($N_{area}$) and phosphorus ($P_{area}$) that are critical to leaf metabolism, including respiration [@meirLeafRespirationTwo2001;@weerasingheCanopyPositionAffects2014] and photosynthetic processes [@scartazzaInvestigatingEuropeanBeech2016; @niinemetsWithincanopyVariationRate2004a; @weerasingheCanopyPositionAffects2014].
Increases in chlorophyll a/b ratios with height reflect the greater light availability in the upper canopy, while greater chlorophyll concentrations at lower heights within the canopy give shade leaves greater PAR absorptance efficiency (Table `r table_leaf_traits`).
Higher photosynthetic rates (per unit leaf area) and more frequent stomatal closure higher (see "Leaf gas exchange and its thermal sensitivity" section, Table `r table_leaf_metabolism`) result in lower intercellular CO~2~ concentrations and higher $\delta^{13}$C concentrations in leaf tissues (Table `r table_leaf_traits`).

@@

## Leaf gas exchange and its thermal sensitivity

@@

\newpage
**Table `r table_leaf_metabolism`.** Summary of observed variation in Leaf gas exchange and its thermal sensitivity across the vertical gradient and/or between sun and shade leaves. Studies listed here were compiled using a systematic review process, as described in Supplementary Information Methods `r appendix_lit_review_methods`.
